What is an Ancestral Recombination Graph?

### 2021-12-17 paper.tex

@@

\section*{The ARG as an encoding of genetic ancestry}

@@

>>>>

% HYW: this is a rough first pass, aiming to be concise. We should flash out the "recombination" aspect of ARGs here a bit, I think

For a long time, the Ancestral Recombination Graph (ARG) has been an object of fascination
to those interested in theoretical evolutionary genetics, as it captures all knowable genetic
history of a collection of individual genomes. Recently advances in our ability to infer ARGs
has led to a resurgence of interest in this field. However, these have also given rise to certain 
terminological confusion, which we aim to clear up in this paper.

One major source of confusion is the use of the term "ARG" to refer to both a backwards-in-time
generative process (Griffiths) and a structure which can be created by that process \citep[e.g.][]{minichiello2006mapping,mathieson2020ancestry}.

@@

We suggest that the process definition should be called the "coalescent with recombination" (CwR), reserving the term ARG to refer to the graph structure itself.

\section*{The ARG as a structure}

@@

% HYW: this is a rough first pass. We should take care not to assume that the tree sequence
% approach is the only encoding

The other use of the the term ARG, and the one for which we suggest the term be reserved, 
is to describe the *structure* resulting from the genetic process of inheritance. This structure, 
like any other graph, consists of *nodes* connected by *edges*. In the context of genetics, the
nodes represent (haploid) genomes, and the edges represent paths of inheritance. Although for
theoretical purposes a graph like this can be created by backwards-in-time simulations using
the CwR, in reality it is created forwards-in-time, as a population evolves.

Since genetic inheritance is unidirectional, the edges in this graph structure are *directed*,
with a "parent" and "child" node. Moreover, there is a strict temporal order, so that children
cannot be their own parents: in other words, there are no cycles in the graph, and it is
technically a form of "directed acyclic graph" (DAG).

% HYW: NB: worded carefully below, as it should also be possible to encode ARGs even if there is not a single fixed coordinate system

There is one key feature which distinguishes an ARG for other graphs. Although a genome
(graph node) may have more than one parent, the particulate nature of genetic inheritance
means that each letter in the genome can only come from a *single* parent. Hence if we
compare the equivalent letters from multiple extant genomes, their paths of inheritance must
form a *tree* (which, incidentally, is why phylogenetics plays such an important role in evolutionary
theory). As we shall argue, the ability to generate these "local trees" from an ARG is of key
importance.

When describing an ARG, the primary issue, from a practical point of view, is how to annotate
the graph with the details of which piece of genome has come from which parent. There
are two possibilities: the annotations can be associated either with the nodes, or with the edges.

\subsection*{Node annotated recombination}

In the original (Griffiths) formulation, nodes which have multiple parents are annotated with a
*breakpoint*, representing a genomic position. The genome lying to the left of that position is then taken to come from one parent, while the genome lying to the right is taken to come from the other parent. This is an elegantly concise encoding of the parental source, and is used, for example in the ARGweaver .arg format. However, it suffers from three principal drawbacks:

1. It assumes only a single breakpoint per node: problematic for representing e.g. gene conversion or 
multiple chromosomes (but this could be circumvented by having multiple breakpoints)
2. Nodes can only have a maximum of 2 parents. At first sight, this seems a reasonable limitation,
but it turns out that it can be helpful to collapse multiple nodes into one, which potentially results
in more than two parents for any given node (see section XXXX)
3. Generating local trees from such an encoding is time consuming (XXX describe *how* time consuming - i,e. scaling properties with number of nodes - and why this is, theoretically)

\subsection*{Edge annotated recombination}

An alternative approach to annotating nodes with the parent of origin of each section of genome is
to annotate each edge with the genomic region which is inherited via that route. This approach
was first introduced by Kelleher et al (cite msprime paper) and now forms the basis of a comprehensive
software library.

% HYW: should discuss the link to biology & mitosis/meiosis: a "node" is strictly a cell - it is shorthand to associate it with an "individual"
